Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-05 Manuscript/manuscript.Rmd
editted manuscript

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt:2016aa], suggesting that the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @flavahan2016insulator]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. Although several useful insights have been made, such as the convergent binding of CTCF [@ong2014ctcf; @ghirlando2016ctcf], [??? Describe other hallmarks of TAD boundaries, with references. And, most importantly, why you are doing your study?]

many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, the class imbalance issue will contribute to small imbalance ratios (defined as the ratio of minority to majority classes[@Orriols-Puig2008]). Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. <!--A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes.--> Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are composite metrics involving both precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while the MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

@@

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [??? REFS]. The use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

@@

Results showed that the imbalance ratios were well below acceptable thresholds for both cell lines, and across the 4 resolutions (Supplementary Table 9 ???SDs will come with more chromosomes). The low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not. While resolution certainly played a role in determining the number of observations included in datasets used in downstream modeling, as it determines the width of genomic bins and larger widths (lower resolutions) created less bins, we found that there was very little variation in imbalance ratios across resolution, with ICC=X and Y for GM12878 and K562 respectively (Supplementary Figure 16 ???ICC values will come with more chromosomes).

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. In evaluating such composite metrics like F1-Score and MCC, we found that for the GM12878 cell line, these models exhibited poor performances across all resolutions (Figure 5A). <!--The average F1-Score and MCC across resolution was found to be 0.04 and 0.07 respectively.-->

We then gradually increased the granularity in the characteristics of the model by assess the model built on OP predictor types, while still not performing any data resampling. Here we found that performances were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). <!--Here, the average F1-Score and MCC was found to be 0.01 and 0.03 respectively.--> Finally, models built on distance-type predictors in conjuction with no resampling, however, were found to exhibit slightly better performances than the other two types<!--, with average F1-Scores and MCCs of 0.08 and 0.10 respectively--> (Figure 5C).

Importantly, when validating these 3 models built on fully imbalanced datasets, the F1-Score and MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 3). <!--The average AUC across resolutions for the OC, OP, and distance type predictors was 0.65, 0.66, and 0.85 respectively, where as the average accuracies were 0.89, 0.89, and 0.90 respectively (Supplementary Table 3).--> Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases.

@@

Comparisons of the effects of the various resampling techniques were then made. When models were built on randomly oversampled (ROS) data, we found a substantial increase in F1-Score and MCC for the GM12878 cell line (Figure 5). The biggest improvement in performance was seen at the 10 kb resolution, with a downward trend as resolutions decreased. Although performances were low at lower resolutions, both F1-Scores and MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest F1-Scores and MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 10). Here, a much sharper negative trend was featured for both F1-Score and MCC and resolution decreased, than in the GM12878 cell line.

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, for all 3 predictor types (Figure 5). Although the differences in performance between the two techniques were marginal, SMOTE resampled data was found to yield the best F1-Scores and MCCs among all classes of models developed and compared in the ensemble framework. As with ROS, distance type predictors yeilded the highest F1-Score and MCC values across resolutions, when applied to either RUS or SMOTE. SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 10).

Moreover, we found that there was a degredation in model performance as data resolution decreased. This was much more pronounced in the K562 cell line. On the GM12878 cell line, there were only marginal differences in performance between 10 kb and 25 kb, followed by a steep decline at 50 kb. Then, both F1-Scores and MCCs showed a momentary increase in performance at 100 kb, the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, starting with highest performance at 10 kb and ending with the worst performance at 100 kb.

These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.  Additionally, when we performed and evaluated the variable importances of random forest models built on one genomic annotation at a time (including the 3 predictor types each), we found that the distance-type consistently out-ranked the OC and OP type predictors in terms of predictive importance, for a majority of the cell line specific annotations, across resolution and cell line (Supplementary Tables 4 & 5). Therefore, it appears that distance-type predictors provided the most informative spatial representation of epigenomic annotations associated with TAD boundaries.

## VARIABLE IMPORTANCES FROM TOP PERFORMING MODELS CONFIRMED BIOLOGICAL FUNCTIONALITY OF KNOWN TRANSCRIPTION FACTOR BINDING SITES

In addition to adressing both the affect of class imbalance and the optimal spatial relationships used when predicting TAD boundaries, we also aimed at identifying the genomic annotations that were most predictive of TAD boundary formation. In doing so, we investigated which genomic annotations yielded the hight predictive importance values for random forest models implemented on the best given set of data-level characteristics determined above (SMOTE re-sampled with distance type predictors), across resolution and cell line. We found that specific transcription factor binding sites (TFBS), including SMC3ab, ZNF143, CTCF, and RAD21, were given consistently high predictive importance in the models built on the GM12878 cell line, across all resolutions (Figure 8). These are known factors associated with the hypothesized loop extrusion model [@fudenberg2016formation; @alipour2012self] and were found to yield consistently high predictive importance on the K562 cell line as well, with similar profiles across resolutions (Supplementary Figure 15). Other annotations known to be associated with gene regulation and affect structural properties of chromatin, such as the YY1  transcription factor, as well DNAse hypersensitive sites, regions of heterochromatin, and the H3k27 trimethylation histone modifications were ranked high on the 10 kb resolution, but were less consistently ranked at lower resolutions for both cell lines. <!--Moreover, the ranking profiles of genomic annotations for both cell lines were found to be more similar between the 10 kb and 25 kb resolutions, as well as between the 50 kb and 100 kb resoluions (this is proved when looking at a clustered heatmap).-->

## MODELS IMPLEMENTED ON TRANSCRIPTION FACTOR BINDING SITES OUTPERFORM OTHER ANNOTATION-SPECIFIC MODELS

Given the prevalence of TFBS as the most consistently top ranked predictive features across resolution and cell line, we ventured to further compare performances of annotation specific models. After applying SMOTE re-sampling with distance type predictors, random forest models built on sets of histone modifications, chromatin state segmentations, and TFBS were compared to a model that included all three types together. For the GM12878 cell line, models built simply on TFBS were found to perform just as well as models that included all 3 sets of annotations according to F1-Score and MCC, consistently across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the lowest F1-Scores and MCC, across resolutions, followed by models built on chromatin state segmentations.

@@

<!-- IN PROGRESS
# DISCUSSION

We developed an ensemble framework to systematically compare model performances for different combinations of class balancing and feature engineering techniques aimed aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations

Although we were not able to draw direct comparisons between the GM12878 and K562 cell lines, given that they did not contain the same list of cell type specific annotations, we did find that there was relative agreement in rankings of the top 5-7 predictors. However, there were notable exclusions from each cell line, in addition to inconsistencies across resolutions

@@

Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.
-->

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt:2016aa], suggesting that the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @flavahan2016insulator; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

The distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Likewise, emerging research has shown that particular binding sites are enriched at TAD boundaries. It has been established that the zinc finger protein, CTCF, is associated with TAD boundaries and is an architectural protein responsible for the organization of the genome within the nucleus [@ghirlando2016ctcf; @ong2014ctcf]. CTCF along with other transcription factors, including SMC1, SMC3, and RAD21 have been shown to colocalize together near TAD boundaries, leading to the recruitment of a cohesin ring [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Other binding proteins have been shown to mediate enhancer-promoter interactions, including the transcription factors YY1 and ZNF143, thereby influencing the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [rao20143d]. However, the growing number of annotated genomic features and the methods used to identify the ones assocaited with TAD boundaries still leaves much to be desired.

Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. In doing so, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. nonboundary regions. This is because, when constructing the data for modeling, a common methodology used to determine the class of the response variable is to bin the linear genome and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome though, this will inevitably lead to severly imbalanced classes.

In a classification setting, when one is attempting to discriminate between regions with (minority class) and without (majority class) TAD boundaries, the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. A better measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthew's Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

<!--
precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while the MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].
-->

@@

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. The use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

Class imbalance was characterized by each dataset's imalance ratio, defined as the ratio of minority to majority classes given by the genomic bins. Minority classes were the set of genomic bins that overlapped with flanked TAD boundaries, while majority classes were the set of genomic bins that did not overlap with a flanked TAD boundary.

@@

Results showed that the imbalance ratios were well below acceptable thresholds for both cell lines, and across the 4 resolutions (Supplementary Table 9 ???SDs will come with more chromosomes). The low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not. While resolution certainly played a role in determining the number of observations included in datasets used in downstream modeling, as it determines the width of genomic bins and larger widths (lower resolutions) created less bins, we found that there was very little variation in imbalance ratios across resolution (Supplementary Figure 16 ???variation will come with additional chrs).

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built data with fully imbalanced classes exhibited poor MCC across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). 
Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 3). Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases.

@@

When models were built on randomly oversampled (ROS) data, we found a substantial increase in MCC for the GM12878 cell line (Figure 5 A-C). The biggest improvement in performance was seen at the 10 kb resolution in each of the 3 predictor types, with a downward trend as resolutions decreased. Although performances were low at lower resolutions, MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest MCCs for ROS data, across all resolutions. Similar results were obtained when applying ROS to the K562 cell line (Supplementary Figure 10). Here, a much sharper negative trend was featured for MCCs as resolution decreased, than in the GM12878 cell line.

Further improvement in performance was found when models were implemented on both randomly undersampled (RUS) and SMOTE resampled data, for all 3 predictor types (Figure 5). Although the differences in performance between the two techniques were marginal, SMOTE resampled data was found to yield the best MCCs among all classes of models developed and compared in the ensemble framework. As with ROS, distance type predictors yielded the highest MCC values across resolutions, when applied to either RUS or SMOTE. SMOTE re-sampled data, with the inclusion of distance-type predictors was also found to be the best set of characteristics when applied to the K562 cell line (Supplementary Figure 10). Moreover, we found that there was a degredation in model performance as data resolution decreased. This was much more pronounced in the K562 cell line. On the GM12878 cell line, there were only marginal differences in performance between 10 kb and 25 kb, followed by a steep decline at 50 kb. Then, the MCCs showed a momentary increase in performance at 100 kb, the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, starting with highest performance at 10 kb and ending with the worst performance at 100 kb. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on distance-type predictors and incorporates both random under- and over-sampling.

<!--
Additionally, when we performed and evaluated the variable importances of random forest models built on one genomic annotation at a time (including the 3 predictor types each), we found that the distance-type consistently out-ranked the OC and OP type predictors in terms of predictive importance, for a majority of the cell line specific annotations, across resolution and cell line (Supplementary Tables 4 & 5). Therefore, it appears that distance-type predictors provided the most informative spatial representation of epigenomic annotations associated with TAD boundaries.
-->

## MODELS IMPLEMENTED ON TRANSCRIPTION FACTOR BINDING SITES, USING OPTIMAL PARAMETERS, OUTPERFORM OTHER ANNOTATION-SPECIFIC MODELS

In order to assess the predictive capabilities of specific types of annotations, we further compared the performances of annotation-specific models. Using the optimal set of parameters, SMOTE re-sampling with distance-type predictors, random forest classifiers built on sets of histone modifications, chromatin state segmentations, and TFBS were compared to a model that included all three types together. For the GM12878 cell line, models built simply on TFBS were found to perform consistently just as well as models that included all 3 sets of annotations, across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the MCC, across resolutions, followed by models built on chromatin state segmentations. This suggests that TAD boundary formation is primarily carried out by transcription factor binding sites.

## PREDICTIVE IMPORTANCES CONFIRMED BIOLOGICAL FUNCTIONALITY OF KNOWN TRANSCRIPTION FACTOR BINDING SITES

As an external method of validation, we also investigated the predictive importance of the annotations that our top performing models were built on. We found that specific transcription factor binding sites (TFBS), including SMC3ab, ZNF143, CTCF, and RAD21, were given consistently high predictive importance in the models built on the GM12878 cell line, across all resolutions (Figure 8). These are known factors associated with the hypothesized loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self] and were found to yield consistently high predictive importance on the K562 cell line as well, with similar profiles across resolutions (Supplementary Figure 15).

@@

# DISCUSSION

A typical problem facing machine learning approaches toward predicting TAD boundary formation is the introduction of highly imbalanced classes. This is because, when constructing the data for modeling, a common methodology used to determine the class of the response variable is to bin the linear genome and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome though, this will inevitably lead to severly imbalanced classes. A workaround to this employed by [@huang2015predicting; @hong2017computational; @bednarz2014supervised] was to randomly sample from the majority class to create balanced classes in the training set. It is unclear how these studies handled

We developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. Additionally, this study provides direct comparisons of 4 different Hi-C data resolutions, and their effects on performances of downstream models. Furthermore, we go on to show the importance of using adequate metric when assessing model performance, especially in the case of highly imbalanced classes.

When constructing the data for modeling, a common methodology used to determine the response variable is to bin the linear genome into equally sized intervals and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome, this will inevitably lead to severly imbalanced classes.

Although we were not able to draw direct comparisons between the GM12878 and K562 cell lines, given that they did not contain the same list of cell type specific annotations, we did find that there was relative agreement in rankings of the top predictive predictors. However, there were notable exclusions from each cell line, in addition to inconsistencies across resolutions

@@

Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.
